Estimating the number of clusters using Cross Validation

### 2015-08-27 cvclust.tex
Add setting 7
add single cluster, multiple dimension case

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}

@@

\subsection{Single Cluster}

@@

\begin{proposition}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

@@

\subsection{Two Clusters}

@@

  \item[6.] Three exponential clusters in $20$ dimensions with different
    variance --- $40$ observations in each cluster are generated from centered
    exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension),
    with $\boldsymbol\mu$ randomly generated from
    $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2,
    1/5$ respectively.
\end{enumerate}
Note that in setting 2--6, all clusters are well-separated, i.e. no
overlapping. In fact, any simulated clusters with minimum distance less than
1 unit was discarded, so there is clear definition of true number of
clusters. The parameters in setting 2--6 are chosen such that about half of
the random realization were discarded. The idea is borrowed from
\cite{tibshirani2001estimating}.

@@

\begin{center}
\footnotesize
\begin{longtable}{lcccccccccccccccc}
\captionsetup{justification=centering}
\caption{\label{table1} Simulation Results}\\
\toprule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endfirsthead
\multicolumn{17}{c}%
  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
\midrule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endhead
\multicolumn{17}{r}{\textit{Continued on next page}} \\
\endfoot
\bottomrule
\multicolumn{17}{l}{\footnotesize
  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
  Setting~1 because they cannot select $k=1$.
}
\endlastfoot
\input{demo/bench/setting1/results.tex}
\midrule
\input{demo/bench/setting2/results.tex}
\midrule
\input{demo/bench/setting3/results.tex}
\midrule
\input{demo/bench/setting4/results.tex}
\midrule
\input{demo/bench/setting5/results.tex}
\midrule
\input{demo/bench/setting6/results.tex}
\end{longtable}
\end{center}

>>>>

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\corr}{corr}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}
\newcommand{\T}{T}

@@

\subsection{Single cluster in two dimensions}

@@

\begin{proposition}\label{prop:1clust-2dim}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

@@

\subsection{Single cluster in higher dimensions}

To generalize the result of the previous section to higher dimensions, 
suppose that $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ is data from a single fold of
Gabriel Cross-Validation, with each $(X, Y)$ pair in $\R^p \times \R^q$ a
draw a mean-zero multivariate normal distribution. To analyze this setting, it
suffices to consider the mean-zero case with $\cov(X) = I_p$.

Suppose that $\cov(Y) = U \Lambda U^\T$ is the eigendecomposition of
$\cov(Y)$. For simplicity, suppose that the leading eigenvalue $\lambda_1$
has unit multiplicity, so that the first principal subspace of $Y$ is a line,
given by the span of the vector $u_1 \in \R^q$. When we apply $k$-means to
$Y_1, \dotsc, Y_n$, the centroids will be close to
$\bmuY_1 =  \sqrt{2\lambda_1 /\pi} u_1$ and
$\bmuY_2 = -\sqrt{2\lambda_1 /\pi} u_1$, with errors of size $\OhP(n^{-1/2})$.

The means for the $X$ variables will be close to
$\bmuX_1 = \E(X \mid u_1^\T Y > 0)$ and
$\bmuX_2 = \E(X \mid u_1^\T Y < 0)$,
with errors of size $\OhP(n^{-1/2})$.  Let
\[
  \begin{bmatrix}
    I_p & \lambda_1^{1/2} \rho \\
    \lambda_1^{1/2} \rho^T & \lambda_1
  \end{bmatrix}
\]
be the covariance matrix of $(X, u_1^\T Y)$,
where $\rho = (\rho_1, \dotsc, \rho_p)$ is such that $\rho_i = \corr(X_i,
u_1^\T Y)$ for $i = 1, \dotsc, p$.
. Then,
\[
  \bmuX_1 = \sqrt{2 \lambda_1 / \pi} \rho,
  \qquad
  \bmuX_2 = -\sqrt{2 \lambda_1 / \pi} \rho;
\]
thus, the classification rule learned from $\{ (X_i, \hGY_i) \}_{i=1}^{n}$
will send observations to class $1$ or $2$ according to whether
$\rho^\T X + \OhP(n^{-1/2})$ is positive or negative.

We have reduced the analysis of cross-validation error to the two-dimensional
case. In particular, following the calculations from the proof of
Prop.~\ref{prop:1clust-2dim}, if $|\corr(\rho^\T X, u_1^\T Y)| < 0.5$, then with
probability tending to one as $m$ and $n$ increase, $\CV(1) < \CV(2)$.
The condition $|\corr(\rho^\T X, u_1^\T)| < 0.5$ reduces to
\[
  \sum_{i=1}^p \rho_i^2 < 0.25.
\]
A sufficient (but not necessary) condition to ensure that this holds is that
the magnitude of the canonical correlation between the $X$ variables and
the $Y$ variables is below $0.5 / \sqrt{p}$.

\subsection{Two clusters}

@@

  \item[6.] Three exponential clusters in $20$ dimensions with different
    variance --- $40$ observations in each cluster are generated from centered
    exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension),
    with $\boldsymbol\mu$ randomly generated from
    $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2,
    1/5$ respectively.

  \item[7.] One cluster in $10$ dimensions with correlated noise -- $200$
    mean-zero multivariate normal observations; the covariance matrix $\Sigma$
    has $\Sigma_{ij} = 0.9$ for $i \neq j$ and $\Sigma_{ii} = 1$ for $i = 1,
    \dotsc, 10$.
\end{enumerate}
Note that in setting 2--6, all clusters are well-separated, i.e. no
overlapping. In fact, any simulated clusters with minimum distance less than
1 unit was discarded, so there is clear definition of true number of
clusters. The parameters in setting 2--6 are chosen such that about half of
the random realization were discarded. The idea is borrowed from
\cite{tibshirani2001estimating}.

@@

\begin{center}
\footnotesize
\begin{longtable}{lcccccccccccccccc}
\captionsetup{justification=centering}
\caption{\label{table1} Simulation Results}\\
\toprule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endfirsthead
\multicolumn{17}{c}%
  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
\midrule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endhead
\multicolumn{17}{r}{\textit{Continued on next page}} \\
\endfoot
\bottomrule
\multicolumn{17}{l}{\footnotesize
  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
  Setting~1 because they cannot select $k=1$.
}
\endlastfoot
\input{demo/bench/setting1/results.tex}
\midrule
\input{demo/bench/setting2/results.tex}
\midrule
\input{demo/bench/setting3/results.tex}
\midrule
\input{demo/bench/setting4/results.tex}
\midrule
\input{demo/bench/setting5/results.tex}
\midrule
\input{demo/bench/setting6/results.tex}
\midrule
\input{demo/bench/setting7/results.tex}
\end{longtable}
\end{center}
